### 2008-12-20 talk/ecoop2009/clibackend.tex
a introduction about jit backends and jit layering, fix an XXX

%\subsection{Promotion on CLI}
%
%Implementing promotion on top of CLI is not straightforward, as it needs to
%patch and modify the already generated code, and this is not possible on .NET.
%
%To solve, we do xxx and yyy etc. etc.
%

\anto{XXX: add an overview section of the backend, else we only talk about
  flexswitches}
\cfbolz{it should maybe also have a nice introductory paragraph how the
interface between the JIT works?}

@@

As already explained, \emph{flexswitch} is one of the key
concepts allowing the JIT compiler generator to produce code which can
be incrementally specialized and compiled at run time.

@@

The picture in Figure~\ref{block-id-fig} shows a graph composed of three methods (for
simplicity, dispatchers are not shown); method ids are in red, whereas
block numbers are in black. 
The graph contains three external links; in particular, note the link
between blocks \texttt{0x00020001} and \texttt{0x00010001} which
connects two blocks implemented by different methods.
\begin{figure}[h]
\begin{center}
\includegraphics[height=6cm]{blockid}
\caption{Method and block ids.\dacom{in the picture ``Main method'' must be
    replaced with ``Primary method'' and in the primary method the block with number 0 should be added}}\label{block-id-fig}
\end{center}
\end{figure}

>>>>

\subsection{JIT layering}

From the implementation point of view, the JIT generator is divided into a
frontend and several backends.  The goal of the frontend is to generate a JIT
compiler which works as described in the previous sections.  Internally, the
frontend represents the compiled code as \emph{flow graphs}, and the role of
the backends is to translate flowgraphs into machine code.

At the moment of writing, three backends have been implemented: one for Intel
\emph{x86} processors, one for \emph{PowerPC} processors, and one for the
\emph{CLI Virtual Machine}.  The latter is special because instead of emitting
code for a real CPU it emits code for a virtual machine: before being
executed, the generated code will be compiled again by .NET's own JIT
compiler.

Thus, when using the CLI backend, we actually have two layers of
JIT-compilation, each layer removing away different kinds of overhead.  By
operating at a higher level, our JIT can potentially do a better job than the
.NET one in some contexts, as our benchmarks demonstrate (see
section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
very good at producing machine code, much more than PyPy's own \emph{x86}
backend for example.  By combining the strengths of both we can get highly
efficient machine code.

As usual, the drawback is that programs that runs for a very short period of
time could run slower with JIT than without, due to the time spent doing the
initial (double) compilation.  It is important to underline that so far we
mostly focused making the JIT able to produce very fast code, without trying
to optimzed the compilation phase itself.

@@

For a large part, implementing the CLI backend is easy and straightforward, as
there is a close corrispondence between most of the operations used by
frontend's flowgraphs and the CLI instructions.  Thus, we will not dig into
details for this part.

However the concept of \emph{flexswitch}, as described in
Section~\ref{sec:implementing-promotion}, does not have any direct equivalent
in the CLI model, and it is hard to implement efficiently.

@@

The picture in Figure~\ref{block-id-fig} shows a graph composed of three methods (for
simplicity, dispatchers are not shown); method ids are in red, whereas
block numbers are in black. 
The graph contains three external links; in particular, note the link
between blocks \texttt{0x00020001} and \texttt{0x00010001} which
connects two blocks implemented by different methods.
\begin{figure}[h]
\begin{center}
\includegraphics[height=6cm]{blockid}
\caption{Method and block ids.}\label{block-id-fig}
\end{center}
\end{figure}
